### 2024-09-10 teaching/math240f24/draft.tex
Lecture 4

\proclaim Proposition 2. Let $p$, $q$, and $r$ be propositional variables. Then
\medskip
\item{i)} $p\wedge 1 \equiv p$ and $p\vee 0 \equiv p$;
\smallskip
\item{ii)} $p \vee 1 \equiv 1$ and $p \wedge 0 \equiv 0$;
\smallskip
\item{iii)} $p \vee p \equiv p$ and $p\wedge p \equiv p$;
\smallskip
\item{iv)} $\neg\neg p \equiv p$;
\smallskip
\item{v)} $p \vee q \equiv q \vee p$ and $p\wedge q \equiv q\wedge p $;
\smallskip
\item{vi)} $p\vee (q\vee r) \equiv (p\vee q) \vee r$ and $p\wedge (q\wedge r) \equiv (p\wedge q)\wedge r)$;
\smallskip
\item{vii)} $p\vee (q\vee r) \equiv (p\vee q) \wedge (p\vee r)$ and
$p\wedge (q\vee r) \equiv (p\wedge q)\vee (p\wedge r)$;
\smallskip
\item{viii)} $p\vee (p\wedge q) \equiv p$ and $p\wedge (p\vee q) \equiv p$; and
\smallskip
\item{ix)} $p\vee \bar p \equiv 1$ and $p\wedge \bar p \equiv 0$.\slug
\medskip

@@

\medskip\boldlabel Conditional and biconditional. We now examine the {\it conditional}
logical relation {\mc IF} $p$ {\mc THEN} $q$. It is denoted by $p\To q$ and its truth table is given by
$$
\vbox{
\tabskip=1em plus.2em minus .2em
\halign{
\hfil$#$\hfil & \hfil$#$\hfil & \hfil$#$\hfil \cr
p & q & p\To q \cr
\noalign{\medskip}
\noalign{\hrule}
\noalign{\medskip}
0 & 0 & 1 \cr
0 & 1 & 1 \cr
1 & 0 & 0 \cr
1 & 1 & 1 \cr
}}.
$$
Within the conditional statement, $p$ is called the {\it antecedent}; this is the assumption. The
statement that is asserted, conditional that the antecendent holds, is called the {\it consequent} $q$.
You can quickly check that the relation $p\To q$ is equivalent to $\neg p \vee q$. This fact is useful
when performing mechanical simplifications.
In English, the statement ``if $p$ then $q$'' asserts a causal relation between $p$ and $q$.
Take a moment and reconcile this idea with the truth table above.
It is normal to get a little tripped up if it's your first time seeing this table.
In the first row, $q$ doesn't even happen, so it might feel weird to say that $p$
has ``caused'' $q$ to happen in this case, and
in the second row, $p$ is not true
and $q$ is true, so it might seem odd that we have set $p\To q$ to true, because there seems to be no relation
between $p$ and $q$.
But it should make sense if you think
of $p$ as a precondition to a promise $q$, and then considering whether the promise is broken. As an example,
suppose your friend says: ``If it snows tomorrow I'll work in Trottier with you.'' If it doesn't snow
and she doesn't pull up, she hasn't technically broken her promise. If it doesn't
snow and she shows up, then she still hasn't broken her promise. The only way she can break her promise is if
it snows and she doesn't come to Trottier; this situation corresponds to the only $0$-row in the truth table.

@@

The last relation is called the {\it biconditional}, and it asserts that variables $p$ and $q$ are logically
equivalent. That is, $p$ happens if and only if $q$ happens. It's truth table
$$
\vbox{
\tabskip=1em plus.2em minus .2em
\halign{
\hfil$#$\hfil & \hfil$#$\hfil & \hfil$#$\hfil \cr
p & q & p\Leftrightarrow q \cr
\noalign{\medskip}
\noalign{\hrule}
\noalign{\medskip}
0 & 0 & 1 \cr
0 & 1 & 0 \cr
1 & 0 & 0 \cr
1 & 1 & 1 \cr
}}
$$
is pretty self-explanatory; in it, $p\Leftrightarrow q$ is true whenever $p$ and $q$ have the same truth value.
\bye
The name ``biconditional'' is suggested by (part of) the following proposition.

@@

This problem is called the {\it boolean satisfiability problem}, often abbreviated {\mc SAT}. One
way of solving it for any given $f$ is to just compute its truth table. We already know the
downside of this approach: if $f$ has $n$ variables, then its truth table will have $2^n$ rows.
Given a few minutes, you are certainly capable of writing down a formula $f$ that has $300$
variables, call them $p_1, \ldots, p_{300}$. The truth table of $f$ will have $2^{300}$ rows,
which is more than the number of atoms in the observable universe. You can learn a lot more
about {\mc SAT} in a higher-level class on computational complexity (e.g., {\mc COMP 360/362}).

@@

\proclaim Proposition. Let $X$ be a finite nonempty set. Let $E$ be the set of all subsets of $X$ with
even cardinality, and let $F$ be the set of all subsets of $X$ with odd cardinality.
Then $|E| = |F|$.

@@

So we can find subset of $\NN$ equipotent with it. It turns out we can also find supersets of $\NN$ with
the same property.

\proclaim Theorem \advthm. We have $|\NN| = |\ZZ|$.

@@

Now we show that $f$ is surjective. Let $k\in \ZZ$. If $k\ge 0$, then consider $n=2k$. We have
$$f(n) = f(2k) = {2k\over 2} = k.$$
If $k<0$, then consider $n=-2k-1$. (Check that this is an element of $\NN$.) Then
$$f(n) = f(-2k-1) = -{-2k-1+1\over 2} = -{-2k\over 2} = k.$$
This shows that $f$ is surjective, so $f$ is in fact bijective and we conclude that $\NN=\ZZ$.\slug

We say that a set $A$ is {\it countably infinite} if there exists a bijection $f:\NN\to A$, that is,
if $|\NN| = |A|$.
A set is said to be {\it countable} if it is either finite or countably infinite.
Otherwise it is called {\it uncountable}.
The previous theorem shows that $\ZZ$ is countably infinite.
If a set $A$ is countable, then there exists an enumeration
$$A = \{a_0, a_1, a_2, \ldots\}.$$
This is useful in the proof of the following theorem.

\proclaim Theorem \advthm. If $A$ and $B$ are countably infinite sets,
then $A\times B$ is also countably infinite.

\noindent{\it Proof sketch.}\enspace
Enumerate $A = \{a_1, a_2, \ldots\}$ and $B = \{b_1, b_2, \ldots\}$. Now given
$(a_i, b_j)\in A\times B$ for some $i,j\in \NN$, we can let
$$f(a_i, b_j) = {(i+j+1)(i+j+2)\over 2} - i - 1.$$
Work out the values of $f(a_i, b_j)$ for small choices of $i$ and $j$, and draw a picture to convince yourself that
$f : A\times B \to\NN$ is a bijection. (Rigorously proving this fact
is quite tedious, so we omit the details.) This means that $f^{-1}$
is a bijection from $\NN$ to $A\times B$, so $A\times B$ is countably infinite.\slug

\proclaim Corollary \advthm. We have $|\ZZ\times \ZZ| = |\NN|$.

@@

One can also show that $|\QQ| = |\NN|$. We will not include the proof, but the idea is that there
is a surjective map from $\ZZ\times\bigl( \ZZ \setminus \{0\}\bigr)\to \QQ$ given by $(m,n)\mapsto m/n$.
On the other hand, there are infinite sets that are not countably infinite. An {\it infinite binary
string} is a function $f:\NN\to \{0,1\}$. (By listing out
$$ f(0), f(1), f(2), \ldots$$
you get an infinite string of $0$s and $1$.)

\proclaim The set $A$ of all infinite binary strings is uncountable.

@@

The set of all infinite binary strings is in bijection with the set of all real numbers in the interval
$[0,1)$. If we're being extra pedantic, we need to forbid an infinite trailing string of all $1$s,
since, e.g., $0.0111\ldots = 0.1000\ldots$ after carrying the $1$s (akin to the fact
that, e.g., $0.0999\ldots = 0.1$). After dealing with this detail, we will have shown that
$\bigl| [0,1]\bigr| \ne |\NN|$.

>>>>

\proclaim Proposition 2. Let $p$, $q$, and $r$ be propositional variables. Then
\medskip
\item{i)} $p\wedge 1 \equiv p$ and $p\vee 0 \equiv p$;
\smallskip
\item{ii)} $p \vee 1 \equiv 1$ and $p \wedge 0 \equiv 0$;
\smallskip
\item{iii)} $p \vee p \equiv p$ and $p\wedge p \equiv p$;
\smallskip
\item{iv)} $\neg\neg p \equiv p$;
\smallskip
\item{v)} $p \vee q \equiv q \vee p$ and $p\wedge q \equiv q\wedge p $;
\smallskip
\item{vi)} $p\vee (q\vee r) \equiv (p\vee q) \vee r$ and $p\wedge (q\wedge r) \equiv (p\wedge q)\wedge r)$;
\smallskip
\item{vii)} $p\vee (q\wedge r) \equiv (p\vee q) \wedge (p\vee r)$ and
$p\wedge (q\vee r) \equiv (p\wedge q)\vee (p\wedge r)$;
\smallskip
\item{viii)} $p\vee (p\wedge q) \equiv p$ and $p\wedge (p\vee q) \equiv p$; and
\smallskip
\item{ix)} $p\vee \bar p \equiv 1$ and $p\wedge \bar p \equiv 0$.\slug
\medskip

@@

\medskip\boldlabel\datestamp{10}{ix} Conditional and biconditional. We now examine the {\it conditional}
logical relation {\mc IF} $p$ {\mc THEN} $q$. It is denoted by $p\To q$ and its truth table is given by
$$
\vbox{
\tabskip=1em plus.2em minus .2em
\halign{
\hfil$#$\hfil & \hfil$#$\hfil & \hfil$#$\hfil \cr
p & q & p\To q \cr
\noalign{\medskip}
\noalign{\hrule}
\noalign{\medskip}
0 & 0 & 1 \cr
0 & 1 & 1 \cr
1 & 0 & 0 \cr
1 & 1 & 1 \cr
}}.
$$
Within the conditional statement, $p$ is called the {\it antecedent}; this is the assumption. The
statement that is asserted, conditional that the antecendent holds, is called the {\it consequent} $q$.
You can quickly check that the relation $p\To q$ is equivalent to $\neg p \vee q$. This fact is useful
when performing mechanical simplifications.
In English, the statement ``if $p$ then $q$'' asserts a causal relation between $p$ and $q$.
Take a moment and reconcile this idea with the truth table above.
It is normal to get a little tripped up if it's your first time seeing this table.
In the first row, $q$ doesn't even happen, so it might feel weird to say that $p$
has ``caused'' $q$ to happen in this case, and
in the second row, $p$ is not true
and $q$ is true, so it might seem odd that we have set $p\To q$ to true, because there seems to be no relation
between $p$ and $q$.
But it should make sense if you think
of $p$ as a precondition to a promise $q$, and then considering whether the promise is broken. As an example,
suppose your friend says: ``If it snows tomorrow I'll work in Trottier with you.'' If it doesn't snow
and she doesn't pull up, she hasn't technically broken her promise. If it doesn't
snow and she shows up, then she still hasn't broken her promise. The only way she can break her promise is if
it snows and she doesn't come to Trottier; this situation corresponds to the only $0$-row in the truth table.

@@

The last relation is called the {\it biconditional}, and it asserts that variables $p$ and $q$ are logically
equivalent. That is, $p$ happens if and only if $q$ happens. It's truth table
$$
\vbox{
\tabskip=1em plus.2em minus .2em
\halign{
\hfil$#$\hfil & \hfil$#$\hfil & \hfil$#$\hfil \cr
p & q & p\Leftrightarrow q \cr
\noalign{\medskip}
\noalign{\hrule}
\noalign{\medskip}
0 & 0 & 1 \cr
0 & 1 & 0 \cr
1 & 0 & 0 \cr
1 & 1 & 1 \cr
}}
$$
is pretty self-explanatory; in it, $p\Leftrightarrow q$ is true whenever $p$ and $q$ have the same truth value.
The name ``biconditional'' is suggested by (part of) the following proposition.

@@

This problem is called the {\it boolean satisfiability problem}, often abbreviated {\mc SAT}. One
way of solving it for any given $f$ is to just compute its truth table. We already know the
downside of this approach: if $f$ has $n$ variables, then its truth table will have $2^n$ rows.
Given a few minutes, you are certainly capable of writing down a formula $f$ that has $300$
variables, call them $p_1, \ldots, p_{300}$. The truth table of $f$ will have $2^{300}$ rows,
which is more than the number of atoms in the observable universe. You can learn a lot more
about {\mc SAT} in a higher-level class on computational complexity (e.g., {\mc COMP 360/362}).
\bye

@@

\proclaim Proposition \advthm. Let $X$ be a finite nonempty set. Let $E$ be the set of all subsets of $X$ with
even cardinality, and let $F$ be the set of all subsets of $X$ with odd cardinality.
Then $|E| = |F|$.

@@

So we can find subset of $\NN$ equipotent with it. It turns out we can also find supersets of $\NN$ with
the same property.
\proclaim Theorem \advthm. We have $|\NN| = |\ZZ|$.

@@

Now we show that $f$ is surjective. Let $k\in \ZZ$. If $k\ge 0$, then consider $n=2k$. We have
$$f(n) = f(2k) = {2k\over 2} = k.$$
If $k<0$, then consider $n=-2k-1$. (Check that this is an element of $\NN$.) Then
$$f(n) = f(-2k-1) = -{-2k-1+1\over 2} = -{-2k\over 2} = k.$$
This shows that $f$ is surjective, so $f$ is in fact bijective and we conclude that $|\NN|=|\ZZ|$.\slug

We say that a set $A$ is {\it countably infinite} if there exists a bijection $f:\NN\to A$, that is,
if $|\NN| = |A|$.
A set is said to be {\it countable} if it is either finite or countably infinite.
Otherwise it is called {\it uncountable}.
The previous theorem shows that $\ZZ$ is countably infinite.

Sometimes it is difficult to come up with a bijection directly. Instead, we would like to find
an injection from $A$ to $B$ (which, in some sense, shows that $|A|\le |B|$), and then an injection
from $B$ to $A$. This is made possible by the following useful theorem, named for Ernst Schr\"oder
and Felix Bernstein, who independently proved it in 1898. The proof is a bit difficult, so it's
technically outside the scope of the course. For fun, you might try to do it as an exercise.

\parenproclaim Theorem {\advthm}. (Schr\"oder--Bernstein theorem). If there exists an injective
function $f:A\to B$ and another injective function $g:B\to A$, then there is a bijection
$h:A\to B$.

\proof We present the proof as a (difficult) exercise. Here is the roadmap.
Call $b\in B$ {\it unattached} if there is no $a\in A$ such that $f(a) = b$.
Let $h:B\to B$ be given by $h(b)  = f\bigl(g(b)\bigr)$. Given $b,b'\in B$, we shall say that
$b$ is a {\it peer} of $b'$ if either $b = b'$ or there exists some $n\in \NN$ such
that
$$ b = \underbrace{h(h(\cdots(h}_{n\ \hbox{\sevenrm times.}}(b'))\cdots))$$
Say that $b\in B$ is a {\it PAE} if it is the peer of an unattached element. (So unattached
elements are automatically PAEs, by setting $n=0$.)
\medskip
\item{a)} Show that if $a\in A$ is such that
$f(a)$ is a PAE, then
there is a unique element $b^*\in B$ such that $g(b^*) = a$, and that
this element is a PAE.
\medskip
\noindent By part (a), if $f(a)$ is a PAE, it makes sense to speak of $g^{-1}(a)$. It is
the element $b^*\in B$ such that $g(b^*) = a$. From here we define
$$ r(a) = \cases{ g^{-1}(a), & if $f(a)$ is a PAE;\cr
f(a), & otherwise.}$$
\medskip
\item{b)} Show that if $b\in B$ is a PAE
then so is $f\bigl(g(b)\bigr)$.
\item{c)} Show that $r$ is surjective. [{\it Hint}: Do a proof by cases. Every $b\in B$ is either
a PAE or it is not a PAE.]
\smallskip
\item{d)} We already proved in part (a) that if $f(a)$ is a PAE, then so is $r(a)$. Prove
the converse of this statement.
\smallskip
\item{e)} Show that $r$ is injective, and therefore bijective. [{\it Hint}:
Assume $r(a_1) = r(a_2)$, and do a proof by cases again. Part (d) will be useful here.]\slug
\medskip

Once again, this proof is outside the scope of the course. Don't lose sleep over it if you can't do it.
Feel free to email me with questions if you get stuck.

Using the Schr\"oder--Bernstein theorem, we now show that the Cartesian product of two countable sets is also
countable.
In the proof, we shall employ the fact that if a set $A$ is countable, then there exists an enumeration
$$A = \{a_0, a_1, a_2, \ldots\}.$$
(If $f$ is the bijection given by the definition, we can set $a_0 = f(0)$, $a_1 = f(1)$, and so on.)

In the following proof, we'll also use the Fundamental Theorem of Arithmetic, which we state now, and prove
later in the course (in the number theory section). It says that
any integer can be factored as a product of primes, a theorem you should have learned in grade school.

\parenproclaim Theorem {\advthm} (Fundamental Theorem of Arithmetic). Every positive
integer $n\ge 2$ can be factored into a product
$${p_1}^{v_1} {p_2}^{v_2}\cdots {p_n}^{v_n},$$
where $n\ge 0$ is an integer and, for $1\le i\le n$, $p_i$ is a prime and $v_i$ is a positive integer.
This factorisation is unique up to the order of the primes.\slug

\edef\thmcartesiancountable{\the\thmcount}
\proclaim Theorem \advthm. If $A$ and $B$ are countably infinite sets,
then $A\times B$ is also countably infinite.

\proof
Enumerate $A = \{a_0, a_2, \ldots\}$ and $B = \{b_0, b_2, \ldots\}$. Now given
$(a_i, b_j)\in A\times B$ for some $i,j\in \NN$, we can let
$$f(a_i, b_j) = 2^i3^j.$$
We show that this defines an injective function. Suppose that $f(a_i, b_j) = f(a_{i'}, f(b_{j'}).$
Then $2^i3^j = 2^{i'} 3^{j'}$, and by the uniqueness of prime factorisations in the Fundamental
Theorem of Arithmetic, we see that $i = i'$ and $j=j'$.

Now we produce an injection $g : \NN\to A\times B$ by simply setting
$$g(n) = (a_n, b_0).$$
It is clear this is injective, since if $g(m) = g(n)$, then $(a_m, b_0) = (a_n, b_0)$, and
that implies that $m=n$.\slug

\edef\corzedtwocountable{\the\thmcount}
\proclaim Corollary \advthm. We have $|\ZZ\times \ZZ| = |\NN|$.

@@

This corollary concerning $\ZZ\times \ZZ$ allows us to prove that $\QQ$ is countable as well.

\proclaim Theorem. The set $\QQ$ of rational numbers is countable.

\proof We define $f:\QQ\to \ZZ\times \ZZ$ as follows. For an element $q\in \QQ$, we let $q = a/b$ the fraction $q$
written in lowest terms (where $a\in \ZZ$ and $b\in \NN\setminus\{0\}$). Then we set $f(q) = (a,b)$.
This is an injective function because if $f(q) = f(q')$, then writing $q = a/b$ and $q' = a'/b'$ in lowest
terms, we have $(a,b) = (a',b')$, so $q = a/b = a'/b' = q'$.

To define an injection $g: \ZZ\times \ZZ\to \QQ$, we recycle the injection we had from the proof
of Theorem~{\thmcartesiancountable}. We already know that $\ZZ$ is countable, so fix an enumeration
$\ZZ = \{a_0, a_1, a_2, \ldots\}$. Then let $g(a_i, a_j) = 2^i 3^j$. The range of $g$ is a subset of $\NN$,
so {\it a fortiori} it is a subset of $\QQ$, and we already showed before that it is an injection.

We have shown that $|\QQ| = |\ZZ\times \ZZ|$, which in turn shows that $|\QQ| = |\NN|$, after
applying Corollary~{\corzedtwocountable}.\slug

So far, we have just given lots of examples of countably infinite sets. Finally, we give an example
of a set that is not countably infinite.

\proclaim Theorem \advthm. The set $A$ of all infinite binary strings is uncountable.

@@

The set of all infinite binary strings is in bijection with the set of all real numbers in the interval
$[0,1)$. If we're being extra pedantic, we need to forbid an infinite trailing string of all $1$s,
since, e.g., $0.0111\ldots = 0.1000\ldots$ after carrying the $1$s (akin to the fact
that, e.g., $0.0999\ldots = 0.1$). After dealing with this detail, one will have shown that
$\bigl| [0,1]\bigr| \ne |\NN|$.

\advsect Relations

\datestamp{10}{} A {\it relation} on a set $X$ is a subset $R\subseteq X\times X$. If $(a,b)\in R$
we write $aRb$ and say ``$a$ is related to $b$.'' Here are some examples.
\medskip
\item{i)} The set $< = \bigl\{(a,b)\in \RR\times\RR : a<b\bigr\}$ is a relation. For instance, we can
write $(2,\pi)\in <$, or $2<\pi$ (which is the more common notation for this relation).
